\Huge Stellar continuum modelling

### 2023-12-06 paper/main.tex
some methods

\newcommand{\project}[1]{\textit{#1}}
\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\vectheta}{\boldsymbol{\theta}}
\newcommand{\vecpsi}{\boldsymbol{\psi}}
\newcommand{\vecW}{\mathbf{W}}
\newcommand{\vecH}{\mathbf{H}}
\newcommand{\vecV}{\mathbf{V}}

% math macros
\newcommand{\unit}[1]{\mathrm{#1}}
\newcommand{\mps}{\unit{m\,s^{-1}}}
\newcommand{\kmps}{\unit{km\,s^{-1}}}

% notes

@@

\begin{abstract}\noindent
Continuum normalization is often a necessary step when analyzing stellar spectra. The best practice is to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but this can be computationally expensive.
We present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear models: a non-negative matrix factorization to approximate line absorption in theoretical (normalized) spectra, and a sines-and-cosines basis for the joint continuum-instrument response.
The non-negative matrix factorization ensures that eigenspectra are strictly additive, and restricts the predicted normalized stellar flux to be at most 1, with the joint continuum-instrument response component to model the rest.
The linearity of both components ensures that inference is stable and fast.
We apply these methods to \project{Sloan Digital Sky Survey} optical and infrared spectra of all evolutionary stages: from M-dwarfs to white dwarfs.
We find good fits to observed spectra, even in situations where the theoretical grid used to approximate line absorption is provably incorrect for some features.
%The stellar absorption model is built from a non-negative matrix factorization of a theoretical continuum-normalized spectra, which ensures it cannot predict spectra more than 1.
%The constraints restrict the stellar absorption to 
We show that we obtain unbiased estimates of the continuum as a function of stellar parameters and signal-to-noise ratios.
\end{abstract}

@@

We will assume a forward model for the data that includes two components: a component to represent continuum-normalized absorption (e.g., line absorption), and a component to represent the smooth continuum.\footnote{Most stellar spectra are not flux-calibrated; the continuum and instrument response enter multiplicatively, but throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.}

We will assume both components to be linear models. This is not a strict requirement, but keeping linearity ensures that  inference is fast and stable, and in practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalisation.\\

% other assumptions
% - NMF is sufficient for representing continuum-normalized spectra
% - that the theoretical models are OK enough. there are ways they can be wrong (eg hydrogen lines) and still work in practice, or there are ways they can be wrong entirely (eg missing molecular band) and not work at all
% - that sines and cosines are sufficiently flexible for modelling the large scale continuum
The problem set up is where the data are a one-dimensional spectrum with $P$ pixels, each with wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_i$ (with $1 \leq i \leq P$). The forward model for these data can be expressed as the element-wise multiplication of the line absorption model $f(\lambda_i; \vecpsi)$ and the continuum-instrument response $g(\lambda_i;\vectheta)$
\begin{align}
    y_i &= f(\lambda_i;\vecpsi)\cdot{}g(\lambda_i;\vectheta) + \mbox{noise}
\end{align}
where the components $f(...)$ and $g(...)$ are described below.\\

% NMF first
The line absorption model $f(\lambda_i;\vecpsi)$ is constructed from a grid of $N$ theoretical spectra each with $D$  continuum-normalized fluxes, which we denote as $\vec{M}$. If the data always have the same wavelength sampling and a constant line spread function, then a good choice is to use a grid where $D = P$. Other situations are described later in this section. With our $\vec{M}$ matrix of continuum-normalized theoretical flux values, we then defined \emph{stellar absorption} $\vecV$ to be
\begin{align}
    \vec{V} = 1 - \vec{M}
\end{align}
such that $\vec{V} \in \left[0, 1\right)$ and is zero when no line absorption exists. This is a `trick' in order to construct a highly constrained and sparse approximation to the matrix $\vecV$ using non-negative matrix factorization (NMF) such that 
\begin{align}
    \vec{V} \approx \vec{W}\vec{H}
\end{align}
where all elements in $\vec{W}$ and $\vec{H}$ are non-negative. 
We could construct an approximation directly to $\vec{M}$, but this would lead to much denser matrices: there are \emph{no} elements of $\vec{M}$ that are exactly 0, but there are \emph{many} elements in $\vec{V}$ that are 0 (e.g., when there is no line absorption).
For this approximation we must chose the number of components to use $C$, which is significantly smaller than both $N$ and $P$. Here $\vec{W}$ is a $N \times C$ matrix that can be thought of as $C$ amplitudes per theoretical spectrum, and $\vec{H}$ is a $C \times P$ matrix of $C$ corresponding eigenspectra. Figure~\ref{fig:nmf} illustrates the NMF procedure and shows some example infrared eigenspectra $\vec{H}$. \\

% Figure NMF

\noindent{}With the matrix $\vec{H}$ we can now define the line absorption function $f(\lambda_i;\vecpsi)$ 
\begin{align}
    f(\lambda_i;\vecpsi) = 1 - \vecpsi\vecH \label{eq:f}
\end{align}
where $\vecpsi \in [0, \infty)$ is a row vector of $C$ elements. $\vecpsi$ is analogous to a single row in $\vecW$: it represents the $C$ amplitudes needed to reconstruct the stellar absorption from the $C$ eigenspectra in $\vecH$. Note that because $\vecpsi$ and $\vecH$ are both restricted to have non-negative elements, Equation~\ref{eq:f} shows that the maximum value that can be predicted by $f(...)$ is 1. This restricts the flexibility of $f(...)$ to only be able to model continuum-normalized flux values, leaving $g(...)$ to represent the joint continuum-instrument response. While negative continuum-normalized flux values are allowed by Equation~\ref{eq:f} (i.e., $\vecpsi\vecH > 1$), this could be readily compensated by a negative continuum model $g(...) < 0$ and such solutions are not favoured in practice.\\

There are many suitable choices for the continuum-instrument response model $g(\lambda_i;\vectheta)$. Here we chose a sine-and-cosine basis because it is a linear representation, and is sufficiently flexible for modelling the joint continuum-instrument response across a variety of spectrographs. The component $g(\lambda_i;\vectheta)$ is expressed as compactly as
\begin{align}
    g(\lambda_i;\vectheta) = \vec{A}(\lambda_i)\vectheta
\end{align}
where $\vec{A}(\lambda_i)$ returns a design matrix where the elements of the $j$th column
\begin{align}
    \vec{A}_{j}(\lambda_i) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{L}\,\lambda_i\right) & \mbox{for $j$ odd} \\[3ex]
                                       \displaystyle\sin\left(\frac{\pi\,j}{L}\,\lambda_i\right) & \mbox{for $j$ even}\end{array}\right. ~,
\end{align}

The design matrix $\vec{A}(\vec{\lambda})$ can be constructed \emph{a priori} before inference begins.

@@

We experimented with choices of initialisation and inference. Initializing from small ($10^{-12}$) values of $\vecpsi$ and a closed-form solution of $\vectheta$ seemed to work well in many scenarios. Since both model components are linear and have closed-form solutions, we did find some success by alternating between solving for $\vecpsi$ and $\vectheta$, but ultimately chose to optimize all parameters $\{\vecpsi,\vectheta\}$ simultaneously.

\subsection{Application to APOGEE spectra}

@@

>>>>

\newcommand{\project}[1]{\textit{#1}}
\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\vectheta}{\boldsymbol{\theta}}
\newcommand{\vecpsi}{\boldsymbol{\psi}}
\newcommand{\vecW}{\mathbf{W}}
\newcommand{\vecH}{\mathbf{H}}
\newcommand{\vecV}{\mathbf{V}}
\newcommand{\hadamard}{\odot}

% math macros
\newcommand{\unit}[1]{\mathrm{#1}}
\newcommand{\mps}{\unit{m\,s^{-1}}}
\newcommand{\kmps}{\unit{km\,s^{-1}}}
\newcommand{\transpose}{^\top}

% notes
\definecolor{tab:blue}{HTML}{1170aa}
\newcommand{\todo}[1]{\textcolor{tab:blue}{#1}}

@@

\begin{abstract}\noindent
Continuum normalization is often a necessary step when analyzing stellar spectra. The best practice is to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but computing stellar spectra can be expensive.
We present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear models: a non-negative matrix factorization to approximate line absorption in theoretical (normalized) spectra, and a sines-and-cosines basis for the joint continuum-instrument response.
The non-negative matrix factorization ensures that eigenspectra are strictly additive, thereby restricting the predicted normalized stellar flux to be at most 1, with the joint continuum-instrument response component to describe the rest.
The linearity of both components ensures that inference is stable and fast.
We apply these methods to \project{Sloan Digital Sky Survey} optical and infrared spectra of all evolutionary stages: from M-dwarfs to white dwarfs.
We find good fits to observed spectra, even in situations where the theoretical grid used to approximate line absorption is provably incorrect for some features.
%The stellar absorption model is built from a non-negative matrix factorization of a theoretical continuum-normalized spectra, which ensures it cannot predict spectra more than 1.
%The constraints restrict the stellar absorption to 
We show that we obtain unbiased estimates of the continuum as a function of stellar parameters and signal-to-noise ratios.
This approach largely eliminates subjectivity in stellar continuum normalization.
\end{abstract}

@@

We will assume a forward model for the data that includes two components: a component to represent continuum-normalized absorption (e.g., line absorption), and a component to represent the smooth continuum.\footnote{Most stellar spectra are not flux-calibrated; the continuum and instrument response enter multiplicatively and require some work to disentangle the two. Throughout this paper when we refer to the continuum, we mean the joint continuum-instrument response.} We will assume both components to be linear models. This is not a strict requirement, but keeping linearity ensures that  inference is fast and stable, and in practice the linear models we construct seem sufficient to model stellar spectra for the purposes of continuum normalisation.\\

% other assumptions
% - NMF is sufficient for representing continuum-normalized spectra
% - that the theoretical models are OK enough. there are ways they can be wrong (eg hydrogen lines) and still work in practice, or there are ways they can be wrong entirely (eg missing molecular band) and not work at all
% - that sines and cosines are sufficiently flexible for modelling the large scale continuum
The problem can be described as having data that are a one-dimensional spectrum with $P$ pixels, each with wavelength $\lambda_i$, flux $y_i$, and flux error $\sigma_i$ (with $1 \leq i \leq P$). The forward model for these data can be expressed as the element-wise multiplication of the line absorption model $f(\lambda_i; \vecpsi)$ and the continuum-instrument response $g(\lambda_i;\vectheta)$
\begin{align}
    y_i &= f(\lambda_i;\vecpsi)\hadamard{}g(\lambda_i;\vectheta) + \mbox{noise}
\end{align}
where $\hadamard$ represents the element-wise (Hadamard) product and the components $f(...)$ and $g(...)$ are described below.\\

% NMF first
The line absorption model $f(\lambda_i;\vecpsi)$ is constructed from a grid of $N$ theoretical spectra each with $D$  continuum-normalized fluxes, which we denote as $\vec{M}$. In the applications that we present, the data are assumed to have the same wavelength sampling and a consistent line spread function, such that we use a grid where $D = P$. Other situations are permitted, but at inference time there is a need to interpolate the line absorption to the $P$ observed pixels. With our $\vec{M}$ matrix of continuum-normalized theoretical flux values, we then defined \emph{stellar absorption} $\vecV$ to be
\begin{align}
    \vec{V} = 1 - \vec{M}
\end{align}
such that $\vec{V} \in \left[0, 1\right)$ and is zero when no line absorption exists. This is a `trick' that allows us to construct a highly constrained and sparse approximation to the matrix $\vecV$ using non-negative matrix factorization (NMF) such that 
\begin{align}
    \vec{V} \approx \vec{W}\vec{H} \label{eq:nmf}
\end{align}
where all elements in $\vec{W}$ and $\vec{H}$ are non-negative. 
We could construct an approximation directly to $\vec{M}$, but this would lead to much denser matrices: there are \emph{no} elements of $\vec{M}$ that are exactly 0 (i.e., total line absorption), but there are \emph{many} elements in $\vec{V}$ that are 0 (i.e., no line absorption). For this approximation we must select the number of components $C$ to use, which is significantly smaller than both the number of theoretical spectra $N$ and the number of pixels $D$ per theoretical spectrum. Here $\vec{W}$ is a $N \times C$ matrix that can be thought of as $C$ amplitudes per theoretical spectrum, and $\vec{H}$ is a $C \times D$ matrix of $C$ corresponding eigenspectra. Figure~\ref{fig:schematic} illustrates the NMF procedure and shows some example infrared eigenspectra $\vec{H}$. \\

\begin{figure*}
    \caption{A schematic illustrating the non-negative matrix factorization procedure, with some example infrared eigenspectra. \label{fig:schematic}}
\end{figure*}

\noindent{}With the matrix $\vec{H}$ we can now define the line absorption function $f(\lambda_i;\vecpsi)$ 
\begin{align}
    f(\lambda_i;\vecpsi) = 1 - \vecpsi\vecH \label{eq:f}
\end{align}
where $\vecpsi \in [0, \infty)$ is a row vector of $C$ elements. $\vecpsi$ is analogous to a single row in $\vecW$: it represents the $C$ amplitudes needed to reconstruct the stellar absorption from the $C$ eigenspectra in $\vecH$. Note that because $\vecpsi$ and $\vecH$ are both restricted to have non-negative elements, Equation~\ref{eq:f} shows that the maximum value that can be predicted by $f(...)$ is 1. This restricts the flexibility of $f(...)$ to only be able to model continuum-normalized flux values, leaving $g(...)$ to represent the joint continuum-instrument response. While negative continuum-normalized flux values are allowed by Equation~\ref{eq:f} (i.e., $\vecpsi\vecH > 1$), this would be readily compensated by a negative continuum model $g(...) < 0$ and in practice such local minima are unfavoured.\\

There are many suitable choices for the continuum-instrument response model $g(\lambda_i;\vectheta)$. Here we chose a sine-and-cosine basis because it is a linear representation, and is sufficiently flexible for modelling the joint continuum-instrument response across a variety of spectrographs. The component $g(\lambda_i;\vectheta)$ is expressed as compactly as
\begin{align}
    g(\lambda_i;\vectheta) = \vec{A}(\lambda_i)\vectheta
\end{align}
where $\vec{A}(\lambda_i)$ returns a design matrix where the elements of the $j$th column are,
\begin{align}
    \vec{A}_{j}(\lambda_i) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{L}\,\lambda_i\right) & \mbox{for $j$ odd} \\[3ex]
                                       \displaystyle\sin\left(\frac{\pi\,j}{L}\,\lambda_i\right) & \mbox{for $j$ even}\end{array}\right. ~,
\end{align}
\noindent{}and the design matrix $\vec{A}(\vec{\lambda})$ can be constructed \emph{a priori} before inference begins.\\

@@

\subsection{Constructing the line absorption model}

The line absorption model $f(\lambda_i;\vecpsi)$ can be easily constructed from a grid of theoretical continuum-normalized spectra. In later sub-sections we discuss the specific choices made for the \emph{APOGEE} and \emph{BOSS} applications, but for now we discuss some general principles and practices that we found successful. The objective function to minimize
\begin{align}
    \chi^2 = ||\vecV - \vecW\vecH||^2
\end{align}
\noindent{}is the Frobenius norm (the squared approximation error). This can be solved through coordinate descent, or through the multiplicative update rules developed by \citet{Lee:2000}:
\begin{align}
    \vecW_{ij}^{\mathrm{(new)}} \leftarrow \vecW_{ij} \frac{\left[\vecV\vecH\transpose\right]_{ij}}{\left[\vecW\vecH\vecH\transpose\right]_{ij}} \quad \phantom{.}\\
    \vecH_{ij}^{\mathrm{(new)}} \leftarrow \vecH_{ij} \frac{\left[\vecW\transpose\vecV\right]_{ij}}{\left[\vecW\transpose\vecW\vecH\right]_{ij}} \quad .
\end{align}

We used the \texttt{scipy.decomposition.NMF} implementation \citep{scipy} with some minor adjustments to minimize the memory requirements (e.g., allowing for lower precision float types). We used no regularization on $\vecW$ or $\vecH$. We found substantial improvements by initializing $\vecH$ with non-negative double singular value decomposition with zeros filled-in with the average of $\vecV$ (the default behavior in the \texttt{scipy} implementation). Initializing with random or small non-negative values produced comparable results but required many more iterations. We stopped the approximation of $\vecW$ and $\vecH$ after 1,000 iterations of multiplicative updates. This took minutes to hours to complete, depending on the size of the theoretical grid $\vecV$.\\

There are no requirements on the number of grid dimensions (e.g., whether or not to include $[\alpha/\mathrm{Fe}]$, $[\mathrm{C/Fe}]$), and no strict requirements (see Section~\ref{sec:discussion}) on the spacing between grid points. The only implicit requirement is that the grid should approximately span the range of stars that you intent to apply the method.\footnote{This is more of a recommendation than a requirement. For example, in practice we found that a grid trained on OBA-type stars was also sufficient to model many white dwarf spectra.} For this reason, we chose to include as many spectra as our memory constraints would allow. If the number of spectra exceeded memory constraints, we would opt to skip every $n$th theoretical spectrum.\\

%Note that the stellar parameters used to compute the theoretical spectra are ignored when solving for $\vecW$ and $\vecH$ in Equation~\ref{eq:nmf}: only the spectra are used. There are correlations between the stellar parameters and the amplitudes $\vecW$, but as we discuss in Section~\ref{sec:discussion}, we avoid trying to use the amplitudes $\vecW$ (or $\vecpsi$) to interpreting stellar parameters.

%\subsection{Inference}

The method can be readily applied to real data once the eigenspectra $\vecH$ have been computed. We experimented with choices of initialisation and inference. Since both components are linear and have closed-form solutions, we did find some success by alternating between solving for $\vecpsi$ and $\vectheta$, but ultimately chose to optimize all parameters $\{\vecpsi,\vectheta\}$ simultaneously. The number of parameters scales as $C + n_\textrm{regions}(2n_\textrm{degree} + 1)$: $C$ amplitudes for $C$ eigenspectra ($\vecpsi$), and $2n_\textrm{degree} + 1$ sine-and-cosine coefficients ($\vectheta$) per chosen continuum region (e.g., per chip). Initializing from small ($10^{-12}$) values of $\vecpsi$ and a closed-form solution of $\vectheta$ (conditioned on small $\vecpsi$) seemed to work well in many scenarios. \\

\subsection{Application to APOGEE spectra}
\label{sec:apogee}

@@

\begin{figure*}
    \caption{The median pixel $\chi^2$ value as a function of \emph{Gaia} $\mathrm{BP} - \mathrm{RP}$ color for main-sequence stars observed with the BOSS spectrograph. \todo{Should expect to see increasing residuals due to TiO bands not captured by the model, and emission lines.}}
\end{figure*}
